Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-08-07 cordillera-cycle.tex
Change Fraser Plateau to Interior Plateau.

% Todo
% change Fraser Plateau to Interior Plateau

@@

During maximum glaciation, both simulations position the main meridional ice
divide over
the western flank of the Rocky Mountains (Fig.~\ref{fig:snapshots}, lower
panels; Fig.~\ref{fig:mis2}). This result appears to contrast with
palaeoglaciological reconstructions for central and southern British Columbia
that inferred the ice divide in a more westerly position,
over the western margin of the Fraser Plateau
\citep{Ryder.etal.1991,
Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}.
These idincate that a latitudinal saddle connected ice dispersal centres in
the Columbia Mountains with the main ice divide \citep{Ryder.etal.1991,
Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal saddle
does indeed feature in our modelling results, however, in an inverse
configuration between the main ice divide over the Columbia Mountains and a
secondary divide over the southern Coast Mountains (Fig.~\ref{fig:mis2}).

@@

Modelled patterns of ice sheet retreat are relatively consistent between the
two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
southern sector of the modelling domain, including the Puget Lowland, the Coast
and Rocky mountains, and the Fraser Plateau, becomes
completely deglaciated by 12\,ka, whereas a significant ice cover remains over
the Skeena, Selwyn, Mackenzie, Wrangell and Saint-Elias mountains in the
northern sector of the modelling domain. After 12\,ka, deglaciation continues
to proceed across the Liard Lowland with a radial ice margin retreat towards
the surrounding mountain ranges, consistent with the regional melt water record
of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
decay by retreating towards the Selwyn and Skeena mountains. The last remnants
of the Cordilleran ice sheet finally disappear from the Skeena Mountains around
10.6\,ka (EPICA) and 9.4\,ka (GRIP).

@@

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelope
\citep[terminology from][]{Kleman.etal.2006}, having been formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
modelled deglacial flow patterns appear consistent between the GRIP and EPICA
simulations. They depict an active ice sheet retreat in the peripheral areas,
followed by stagnant ice decay in some of the interior regions. Several parts
of the modelling domain do not experience any basal sliding throughout the
deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
includes parts of the Fraser Plateau, major
portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
the Mackenzie Mountains.

@@

However, on the Fraser Plateau, both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
results could be that the Fraser Plateau lineation system predates
deglaciation, as perhaps indicated by some eskers that appear incompatible with
these glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Alternatively,
local geothermal heat associated with volcanic activity on the Fraser Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.

The modelled deglaciation of the Fraser Plateau
consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:lastflow}) positioned
in-between deglaciated mountain ranges
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}.

@@

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains (SMKM),
           the Skeena Mountains (SM), the Coast and Rocky
           Mountains (CRM) and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Fraser Plateau (FP) and the
           Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.}
  \label{fig:locmap}
\end{figure}

@@

% fig:lastflow
\begin{figure*}
  \includegraphics{lastflow}
  \caption{Modelled of the deglacial basal ice velocities. Hatches
           indicate areas that remain non-sliding throughout deglaciation
           (22.0--8.0\,ka), notably including the Fraser Plateau (FP).
           Note the concentric patterns of deglacial flow in the Liard
           Lowland (LL).
           Sliding grid cells were distinguished from non-sliding grid cells
           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.
           Dashed segments indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:lastflow}
\end{figure*}

>>>>

@@

During maximum glaciation, both simulations position the main meridional ice
divide over
the western flank of the Rocky Mountains (Fig.~\ref{fig:snapshots}, lower
panels; Fig.~\ref{fig:mis2}). This result appears to contrast with
palaeoglaciological reconstructions for central and southern British Columbia
that inferred the ice divide in a more westerly position,
over the western margin of the Interior Plateau \citep{Ryder.etal.1991,
Stumpf.etal.2000, Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}.
These idincate that a latitudinal saddle connected ice dispersal centres in
the Columbia Mountains with the main ice divide \citep{Ryder.etal.1991,
Kleman.etal.2010, Clague.Ward.2011, Margold.etal.2013a}. A latitudinal saddle
does indeed feature in our modelling results, however, in an inverse
configuration between the main ice divide over the Columbia Mountains and a
secondary divide over the southern Coast Mountains (Fig.~\ref{fig:mis2}).

@@

Modelled patterns of ice sheet retreat are relatively consistent between the
two simulations (Figs.~\ref{fig:deglacshots} and.~\ref{fig:deglac}). The
southern sector of the modelling domain, including the Puget Lowland, the Coast
and Rocky mountains, and the Interior Plateau of British Columbia, becomes
completely deglaciated by 12\,ka, whereas a significant ice cover remains over
the Skeena, Selwyn, Mackenzie, Wrangell and Saint-Elias mountains in the
northern sector of the modelling domain. After 12\,ka, deglaciation continues
to proceed across the Liard Lowland with a radial ice margin retreat towards
the surrounding mountain ranges, consistent with the regional melt water record
of the last deglaciation \citep{Margold.etal.2013}. Remaining ice continues to
decay by retreating towards the Selwyn and Skeena mountains. The last remnants
of the Cordilleran ice sheet finally disappear from the Skeena Mountains around
10.6\,ka (EPICA) and 9.4\,ka (GRIP).

@@

Because a general tenant in glacial geomorphology is that the majority of
landforms (lineations and eskers) are part of the deglacial envelope
\citep[terminology from][]{Kleman.etal.2006}, having been formed
close inside the retreating margin of ice sheets \citep{Boulton.Clark.1990,
Kleman.etal.1997, Kleman.etal.2010}, we present maps of basal flow directions
immediately preceding deglaciation or at the time of cessation of sliding
inside a cold-based retreating margin (Fig.~\ref{fig:lastflow}). The
modelled deglacial flow patterns appear consistent between the GRIP and EPICA
simulations. They depict an active ice sheet retreat in the peripheral areas,
followed by stagnant ice decay in some of the interior regions. Several parts
of the modelling domain do not experience any basal sliding throughout the
deglaciation phase (Fig.~\ref{fig:lastflow}, hatched areas). This notably
includes parts of the Interior Plateau in British Columbia, major
portions of the Alaskan sector of the ice sheet, and a tortuous ribbon running
from the Northern Rocky Mountains over the Skeena and Selwyn Mountains and into
the Mackenzie Mountains.

@@

However, on the Interior Plateau of south-central British Columbia, both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
results could be that the Interior Plateau lineation system predates
deglaciation, as perhaps indicated by some eskers that appear incompatible with
these glacial lineations \citep[Fig.~9]{Margold.etal.2013a}. Alternatively,
local geothermal heat associated with volcanic activity on the Interior Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.

The modelled deglaciation of the Interior Plateau of British Columbia
consists of a rapid northwards retreat (Fig.~\ref{fig:deglac}) of
southwards-flowing non-sliding ice lobes (Fig.~\ref{fig:lastflow}) positioned
in-between deglaciated mountain ranges
(Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}). This result
appears compatible with the prevailing conceptual model of deglaciation of
central British Columbia, in which mountain ranges emerge from the ice before
the plateau \citep[Fig.~7]{Fulton.1991}.

@@

% fig:locmap
\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} (21.4\,cal\,ka)
           \citep{Dyke.2004}. The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
           (SMKM), the Skeena Mountains (SM), the Coast and Rocky
           Mountains (CRM) and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Interior Plateau of British
           Columbia (IP) and the Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.}
  \label{fig:locmap}
\end{figure}

@@

% fig:lastflow
\begin{figure*}
  \includegraphics{lastflow}
  \caption{Modelled of the deglacial basal ice velocities. Hatches
           indicate areas that remain non-sliding throughout deglaciation
           (22.0--8.0\,ka), notably including the Interior Plateau (IP).
           Note the concentric patterns of deglacial flow in the Liard
           Lowland (LL).
           Sliding grid cells were distinguished from non-sliding grid cells
           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.
           Dashed segments indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:lastflow}
\end{figure*}
